Linear Vlasov Theory in the Shearing Sheet Approximation \\ with Application to the Magneto-Rotational Instability

### 2015-01-02 paper.tex
Fixed typos and made some minor cosmetic changes

The tidal potential $\psi$ is the sum of the centrifugal and the gravitational
potential. In the shearing sheet approximation \citep{Goldreich1965}, this is
given by
\begin{equation}
  \label{eq:tidal-potential}
  \psi = -q\Omega^2 x^2 + \frac{1}{2}\vfreq^2 z^2.
\end{equation}
The angular velocity $\Omega$ is determined from the radial force balance. We
assume that in equilibrium, the magnetic field is frozen into a bulk flow
$\vec{u}$ common to all species ($\vec{E}+\vec{u}\times\vec{B}=0$). If the
gravitational potential is due to a point mass and radial pressure support is
negligible, then $q=3/2$ (Keplerian rotation) and the vertical frequency
$\nu=\Omega$.

@@

Inspection of \cref{eq:hamiltonian} immediately reveals that $y$ is a cyclic
coordinate. A second integral of motion (in addition to the energy integral)
is thus given by the conjugate momentum $p_y$. We will henceforth refer to
$p_y=v_y+S_z{}x$ as the (canonical) angular momentum
\citep[cf.][]{Wisdom1988}. For a given $p_y$, the dynamics may be described in
the reduced phase space $(x,z,p_x,p_z$). A particle traveling in the mid-plane
at the local shear velocity \cref{eq:shear-flow} corresponds to a stationary
point in this reduced phase space. To see this, we note that the first
variation of $\mathcal{H}$ is given by
\begin{equation}
  \delta\mathcal{H} = v_x\delta p_x + v_z\delta p_z
  + [S_y v_z - S_z(v_y + q\Omega x)]\delta x + \vfreq^2 z\delta z.
\end{equation}
A stationary point is reached if all coefficients in this expression vanish
identically. This is the case if and only if
\begin{equation}
  z = 0 \quad\textrm{and}\quad v_x = v_y + q\Omega x = 0.
\end{equation}
These orbits are the local representation of global circular orbits (see
\cref{app:global}). In the following this will be implied and we will refer to
them simply as circular orbits.

@@

Throughout much of the subsequent analysis we will make no assumptions about
the specific form of $f(\vc_\perp,\vc_\parallel)$. Starting with section
\cref{sec:schwarzschild-response} we will, however, assume that the
distribution function is given by a Maxwellian in $\vvc$-space, to wit
\begin{equation}
  \label{em:schwarzschild}
  f(\mathcal{K}) =
  \frac{n\exp(-\mathcal{K}/\vt^2)}{{(2\pi)}^{3/2}\vt^3\sqrt{1-\Delta}},
\end{equation}
where $\vt=\mathrm{const}$ is the thermal velocity. In the guiding center
limit ($e/m\to\infty$), this reduces to a drifting Maxwellian in
$\vec{v}$-space \citep[as used in e.g.][]{Quataert2002}. Indeed, as
$e/m\to\infty$, the tidally induced anisotropy $\Delta\sim{}m/e$ disappears.
From the definitions given in
\cref{eq:gyrotropic-velocity,eq:gyro-frequency-vector} it then follows that
$\vvc\approx\vec{v}+q\Omega{}x\ey$ and $\vomg\approx\vomc$. In the limit, any
distribution function of the form $f(\vc_\perp,\vc_\parallel)$ thus becomes
drifting gyrotropic distribution in $\vec{v}$-space, with the anisotropy being
due to the magnetic field only.

@@

We note that \cref{eq:vlasov-relative,eq:faraday-relative} are exact. No
approximations have been made so far. We also note that the only explicit
$x$-dependence in \cref{eq:vlasov-relative,eq:faraday-relative} is multiplying
$\partial/\partial{}y$. Provided the dynamics are axisymmetric, we may thus
take the distribution functions $f_s$, the magnetic field $\vec{B}$, and the
\emph{relative} electric field $\tvec{E}$ to be periodic in $x$.\footnote{More
  generally, if $\partial/\partial{}y\ne0$, then each field appearing in
  \cref{eq:vlasov-relative,eq:faraday-relative} may be written as a sum of
  so-called shearing waves \citep{Thomson1887,Goldreich1965}. In the context
  of computer simulations, the relative electric field $\tvec{E}$ as well as
  $\vec{B}$ and $f_s$ may be taken to be shearing periodic
  \citep{Lees1972,Hawley1995}.}

@@

From now on we will assume no stratification ($\vfreq=0$) and axisymmetric
disturbances ($\partial/\partial{}y=0$). The dynamical equations then have
constant coefficients and we may assume that all disturbances depend on
position $\vec{r}$ and time $t$ as $\exp(ik_x{}x+ik_z{}z-i\omega{}t)$. For
such disturbances, Faraday's law \labelcref{eq:faraday-relative} is easily
solved for $\delta\vec{B}$, to wit
\begin{equation}
  \label{eq:linearized-faraday}
  \delta\vec{B} = \frac{\vec{k}}{\omega}\times\left[
    \left(\mathbf{1} - \frac{q\Omega}{i\omega}\ex\ey\right)
    \cdot\delta\tilde{\vec{E}}
  \right].
\end{equation}
Here, the second term in parenthesis arises from the stretching term in
\cref{eq:faraday-relative}. Substituting \cref{eq:linearized-faraday} into
\cref{eq:formal-solution}, taking the first moment, multiplying by $\qs{}n_s$,
and summing over species yields the perturbed current density
\begin{equation}
  \label{eq:perturbed-current}
  \delta\vec{J}(\vec{r},t)
  = -\frac{1}{i\omega}\sums\frac{\qs^2n_s}{m_s}
  \int d^3v\,\int_{-\infty}^t dt'
  \tvec{v}\pder{\hat{f}_s}{\tvec{v}'}\cdot\big[
  i(\omega - \vec{k}\cdot\tvec{v}')\mathbf{1}
  + i\vec{k}\tvec{v}' + q\Omega\ex\ey
  \big]\cdot\left(\mathbf{1} - \frac{q\Omega}{i\omega}\ex\ey\right)
  \cdot\delta\tvec{E}(\vec{r}',t'),
\end{equation}
where $\hat{f}_s=f_s/n_s$ is the reduced distribution function.

@@

We are now in a position to write down the general dispersion relation of a
charge-neutral collisionless plasma in the shearing sheet approximation.
Inserting the solution of Faraday's law given in \cref{eq:linearized-faraday}
into Ampère's law $\mu_0\delta\vec{J}=i\vec{k}\times\delta\vec{B}$ (neglecting
the displacement current), and eliminating the perturbed current using
\cref{eq:conductive-relation} yields
\begin{equation}
  \label{eq:dispersion-relation}
  \mathbf{D}\cdot\left(
    \mathbf{1} - \frac{q\Omega}{i\omega}\ex\ey
  \right)\cdot\delta\tvec{E} = 0,
\end{equation}
where the dispersion tensor
\begin{equation}
  \label{eq:dispersion-tensor}
  \mathbf{D} =
  (k^2\mathbf{1} - \vec{k}\vec{k} - i\omega\mu_0\vec{\sigma})\va^2
  - 2q\Omega^2\sums\frac{n_s m_s}{\rho}
  \frac{\omcs/b_z}{\omcs b_z + 2\Omega}\ey\ey.
\end{equation}
Here, $\rho=\sums{}n_s{}m_s$ is the mass density, $\va^2=B^2/(\mu_0\rho)$ is
the square of the Alfvén speed, and we have used the charge-neutrality
condition to rewrite the last term on the right hand side.

@@

Here we consider the stability of an initially unmagnetized cold plasma in the
shearing sheet. The dispersion relation for this case is obtained from the
general expression \cref{eq:dispersion-tensor} together with
\cref{eq:conductivity} by substituting \cref{eq:cold-response} and then
letting the magnetic field strength tend to zero. The dispersion tensor in
this limit reduces to
\begin{equation}
  \label{eq:unmagnetized-dispersion-tensor}
  \va^{-2}\ell^2\mathbf{D} =
  (k^2\mathbf{1} - \vec{k}\vec{k} - i\omega\mu_0\vec{\sigma})\ell^2
  + q/2\ey\ey,
\end{equation}
where $\ell$ is the mean inertial length, defined through
\begin{equation}
  \ell^{-2} = \mu_0\sums\frac{\qs^2n_s}{m_s}.
\end{equation}
The limiting form of the cold plasma conductivity tensor is
\begin{equation}
  \label{eq:zero-mag-conductivity}
  -i\omega\mu_0\vec{\sigma} =
  \ell^{-2}\mathbf{Q}\cdot\mathbf{\Lambda}\cdot\mathbf{Q},
\end{equation}
where the anisotropy tensor is given in \cref{eq:anisotropy-tensor} with
$\Delta_s=q/2$ and the cold plasma response tensor for $B=0$ is
\begin{equation}
  \label{eq:zero-mag-response}
  \mathbf{\Lambda} = \ez\ez
  + \frac{1}{2}\sum_{\pm}\frac{\omega}{\omega\pm\kappa}
  (\ex\pm i\ey)(\ex\mp i\ey).
\end{equation}
With this, setting the determinant of \cref{eq:unmagnetized-dispersion-tensor}
equal to zero yields the dispersion relation
\begin{equation}
  \frac{\omega^2{(1 + k^2\ell^2)}^2 - \kappa^2 k_z^2\ell^2(q/2 + k^2\ell^2)}
  {\omega^2 - \kappa^2} = 0.
\end{equation}
Assuming $\omega^2\ne\kappa^2$, the solution is
\begin{equation}
  \omega^2 =
  \frac{\kappa^2k_z^2\ell^2(q/2 + k^2\ell^2)}{{(1 + k^2\ell^2)}^2}.
\end{equation}
This is manifestly positive as long as $q>0$. Thus there is no instability as
long as the orbital frequency decreases radially outward.

@@

In most plasmas, electrons are much lighter than ions, and so the mass ratio
$m_e/m_i\ll1$. If we assume that the electrons are also much colder than the
ions ($T_e/T_i\ll1$), then the electrons can be modeled as a cold massless
fluid. Expanding the cold electron conductivity tensor in the mass ratio
yields
\begin{equation}
  \label{eq:cold-electron-conductivity}
  \vec{\sigma}_e = -\frac{1}{i\omega}\frac{e^2n_e}{m_e}\vec{b}\vec{b}
  + \frac{en_e}{B}\vec{b}\times\mathbf{1},
\end{equation}
where we have dropped terms of order $m_e/m_i$ and higher. The first term
evidently diverges as the mass ratio goes to zero. Because this term is
proportional to $\vec{b}\vec{b}$, the dispersion tensor becomes
block-diagonal.\footnote{We note that since
  $\mathbf{D}=-i\omega\mu_0\vec{\sigma}_e\va^2+\ldots$, it follows from the
  dispersion relation given in \cref{eq:dispersion-relation} that the parallel
  component of the (relative) electric field must be at least first order in
  $m_e/m_i$. The parallel current density would otherwise diverge, a result
  that is consistent with guiding center theory \citep[e.g.][]{Grad1961}.} In
the cold and massless electron limit, the dispersion relation thus factorizes
into $\delta{}\tilde{E}_\parallel=0$ and $\det\mathbf{D}_\perp=0$, where the
perpendicular dispersion tensor
\begin{equation}
  \label{eq:perpendicular-dispersion}
  \mathbf{D}_\perp =
  {(k^2\mathbf{1} - \vec{k}\vec{k} - i\omega\mu_0\vec{\sigma}_i)}_\perp\va^2
  - i\omega\ome\vec{b}\times\mathbf{1}
  - 2q\Omega^2\sums\frac{n_s{}m_s}{\rho}\frac{\omcs b_z}{\omcs b_z + 2\Omega}
  (\vec{b}\times\ex)(\vec{b}\times\ex).
\end{equation}
We note that throughout this section, the labels $\perp$ and $\parallel$ refer
to the direction of $\vec{b}$, \emph{not} to the gyro-frequency vector $\vomg$
as in \cref{sec:steady-state,sec:linear-theory}. In
\cref{eq:perpendicular-dispersion}, $\vec{\sigma}_i$ is the (multi-component)
ion conductivity tensor and
\begin{equation}
  \ome = \frac{en_e B}{\rho} = \sums\frac{n_s m_s}{\rho}\omcs
\end{equation}
is the (mass density weighted) mean ion cyclotron frequency.

@@

In the remainder of this section we will assume that there is only one ion
species. The perpendicular dispersion tensor in this case is given by
\begin{equation}
  \label{eq:perpendicular-dispersion-single}
  \mathbf{D}_\perp =
  \va^2{(k^2\mathbf{1} - \vec{k}\vec{k})}_\perp
  + \omc^2{(\mathbf{Q}\cdot\mathbf{\Lambda}\cdot\mathbf{Q})}_\perp
  - i\omega\omc\vec{b}\times\mathbf{1}
  - 2q\Omega^2\omc b_z/S_z
  (\vec{b}\times\ex)(\vec{b}\times\ex).
\end{equation}
We focus on the Vlasov-fluid limit in this section to make contact with the
existing kinetic theory calculations of the MRI, all of which have assumed
cold electrons. We stress, however, that the results derived in the previous
section are fully general and can be used to quantify the effects of kinetic
electrons on the MRI\@. The cold electron approximation used here and in
previous studies formally requires that $\zeta_n\rightarrow\infty$ for all $n$
(\cref{sec:cold}). However,
$\zeta_0=\omega/|k_\parallel|v_{t,e}\sim{}\va/v_{t,e}$ for the MRI (taking
$\omega\sim{}k_z{}v_A$). Thus $\zeta_0\gg{}1$ requires
$T_e/T_i\ll(m_e/m_i)\beta_i^{-1}$. This is in general \emph{not} satisfied so
that the cold electron approximation is not formally applicable when studying
the MRI in hot accretion flows. In future work, we will study the effect of
(hot) kinetic electrons on the MRI in collisionless plasmas.

@@

In the limit $e/m\to\infty$ we should be able to recover the results of
\citet{Quataert2002} who analyzed the stability of a differentially rotating
plasma in the shearing sheet approximation using the so-called kinetic MHD
equations \citep{Grad1961,Kulsrud1983,Hazeltine2004}. We start by expanding
the response tensor in powers of $m/e$. Keeping mostly leading order terms,
this yields
\begin{equation}
  \mathbf{\Lambda} \approx
  -\frac{\omega^2}{\omc^2}(\mathbf{1} - \vec{e}_3\vec{e}_3)
  + \frac{i\omega}{\omc}\left[1 - \frac{(4-q)\Omega b_z}{2\omc}\right]
  \vec{e}_3\times\mathbf{1}
  - \frac{2(W-1)k_\perp^2\vt^2}{\omc^2}\vec{e}_2\vec{e}_2
  + \frac{i\zeta_0 Wk_\perp\vt}{\omc}\vec{e}_1\times\mathbf{1}
  - \zeta_0^2 W\vec{e}_3\vec{e}_3.
\end{equation}
The argument of the $W$-function is $\zeta_0=\omega/|k_\parallel|\vt$.
Expanding the anisotropy tensor $\mathbf{Q}$ yields
\begin{equation}
  \mathbf{Q} \approx \mathbf{1} - \frac{q\Omega}{2\omc b_z}
  \left[1 - \frac{(8-q)\Omega}{4\omc b_z}\right]\ey\ey.
\end{equation}

@@

In this section we study modes whose wave vector is aligned with the magnetic
field, which in turn is aligned with the rotation axis. The ion response
tensor in this case is given by
\begin{equation}
  \label{eq:parallel-response}
  \mathbf{\Lambda} =
  \frac{1}{2}\sum_\pm\Lambda^{}_\pm
  (\vec{e}_1\pm i\vec{e}_2)(\vec{e}_1\mp i\vec{e}_2)
  - \frac{\omega^2}{k_z^2\vt^2} W\left(\frac{\omega}{|k_z|\vt}\right)
  \vec{e}_3\vec{e}_3,
\end{equation}
with perpendicular components
\begin{equation}
  \label{eq:circular-components}
  \Lambda_\pm = \frac{\omega}{\omega\pm\omg}
  \left[1 - W\left(\frac{\omega\pm\omg}{|k_z|\vt}\right)\right].
\end{equation}
We note that $\ex\ne\vec{e}_1$ and $\ey\ne\vec{e}_2$ in general even for
parallel modes. This is because the orthonormal basis
$(\vec{e}_1,\vec{e}_2,\vec{e}_3)$ with $\vec{e}_3=\vomg/\omg$ is taken to be
right-handed (see \cref{sec:plasma-response}), and whether or not the
gyro-frequency vector $\vomg$ defined in \cref{eq:gyro-frequency-vector} is
aligned with the angular velocity vector $\vec{\Omega}=\Omega\ez$ depends on
the sign of $S_z=\omc{}b_z+2\Omega$. In this context we also note that the
gyration frequency $\omg>0$ throughout this text.\footnote{Unless of course
  $\omg^2<0$, in which case much of our analysis is invalidated, see the
  discussion in \cref{sec:delta}.} The cyclotron frequency $\omc=eB/m$ can
have either sign, but this depends only on the sign of the charge, \emph{not}
on direction of the magnetic field.

@@

We seek an axisymmetric equilibrium with a purely rotational bulk flow common
to all plasma species. In a cylindrical coordinate system $(r,\varphi,z)$,
this bulk flow may be written as
\begin{equation}
  \vec{u} = r\Omega\nabla\varphi,
\end{equation}
where $\Omega(r,z)$ is the orbital frequency. Following
\citet{Kaufman1972,Cary1983} we write the magnetic vector potential as
\begin{equation}
  \label{eq:global-bfield}
  \vec{B} =
  \nabla w(\psi)\times\nabla\vartheta - \nabla\psi\times\nabla\varphi,
\end{equation}
where $\psi(r,z)$ is the poloidal flux function. The poloidal angle
$\vartheta(r,z)$ may be chosen to increase clockwise from $-\pi$ to $\pi$
along poloidal field lines, which are isocontours of $\psi$. Note that the
first term on the right hand side of \cref{eq:global-bfield} represents the
toroidal field. The second term is the usual parametrization of an
axisymmetric poloidal field. The magnetic field given in
\cref{eq:global-bfield} derives from the axisymmetric vector potential
\begin{equation}
  \vec{A} = w(\psi)\nabla\vartheta - \psi\nabla\varphi.
\end{equation}

>>>>

The tidal potential $\psi$ is the sum of the centrifugal potential and the
gravitational potential. In the shearing sheet approximation
\citep{Goldreich1965}, this is given by
\begin{equation}
  \label{eq:tidal-potential}
  \psi = -q\Omega^2 x^2 + \frac{1}{2}\vfreq^2 z^2.
\end{equation}
The angular velocity $\Omega$ is determined from radial force balance. We
assume that in equilibrium, the magnetic field is frozen into a bulk flow
$\vec{u}$ common to all species ($\vec{E}+\vec{u}\times\vec{B}=0$). If the
gravitational potential is due to a point mass and radial pressure support is
negligible, then $q=3/2$ (Keplerian rotation) and the vertical frequency
$\nu=\Omega$.

@@

Inspection of \cref{eq:hamiltonian} immediately reveals that $y$ is a cyclic
coordinate. A second integral of motion (in addition to the energy integral)
is thus given by the conjugate momentum $p_y$. We will henceforth refer to
$p_y=v_y+S_z{}x$ as the (canonical) angular momentum
\citep[cf.][]{Wisdom1988}. For a given $p_y$, the dynamics may be described in
the reduced phase space $(x,z,p_x,p_z$). A particle traveling in the mid-plane
at the local shear velocity \labelcref{eq:shear-flow} corresponds to a
stationary point in this reduced phase space. To see this, we note that the
first variation of $\mathcal{H}$ is given by
\begin{equation}
  \delta\mathcal{H} = v_x\delta p_x + v_z\delta p_z
  + [S_y v_z - S_z(v_y + q\Omega x)]\delta x + \vfreq^2 z\delta z.
\end{equation}
A stationary point is reached if all coefficients in this expression vanish
identically. This is the case if and only if
\begin{equation}
  z = 0 \quad\textrm{and}\quad v_x = v_y + q\Omega x = 0.
\end{equation}
These orbits are the local representation of global circular orbits (see
\cref{app:global}). In the following this will be implied and we will refer to
them simply as circular orbits.

@@

Throughout much of the subsequent analysis we will make no assumptions about
the specific form of $f(\vc_\perp,\vc_\parallel)$. Starting with section
\cref{sec:schwarzschild-response} we will, however, assume that the
distribution function is given by a Maxwellian in $\vvc$-space, to wit
\begin{equation}
  \label{em:schwarzschild}
  f(\mathcal{K}) =
  \frac{n\exp(-\mathcal{K}/\vt^2)}{{(2\pi)}^{3/2}\vt^3\sqrt{1-\Delta}},
\end{equation}
where $\vt=\mathrm{const}$ is the thermal velocity. In the guiding center
limit ($e/m\to\infty$), this reduces to a drifting Maxwellian in
$\vec{v}$-space \citep[as used in e.g.][]{Quataert2002}. Indeed, as
$e/m\to\infty$, the tidally induced anisotropy $\Delta\sim{}m/e$ disappears.
From the definitions given in
\cref{eq:gyrotropic-velocity,eq:gyro-frequency-vector} it then follows that
$\vvc\approx\vec{v}+q\Omega{}x\ey$ and $\vomg\approx\vomc$. In the limit, any
distribution function of the form $f(\vc_\perp,\vc_\parallel)$ thus becomes a
drifting gyrotropic distribution in $\vec{v}$-space, with the anisotropy being
due to the magnetic field only.

@@

We note that \cref{eq:vlasov-relative,eq:faraday-relative} are exact. No
approximations have been made so far. We also note that the only explicit
$x$-dependence in \cref{eq:vlasov-relative,eq:faraday-relative} is multiplying
$\partial/\partial{}y$. Provided the dynamics are axisymmetric, we may thus
take the distribution function $f_s$, the magnetic field $\vec{B}$, and the
\emph{relative} electric field $\tvec{E}$ to be periodic in $x$.\footnote{More
  generally, if $\partial/\partial{}y\ne0$, then each field appearing in
  \cref{eq:vlasov-relative,eq:faraday-relative} may be written as a sum of
  so-called shearing waves \citep{Thomson1887,Goldreich1965}. In the context
  of computer simulations, the relative electric field $\tvec{E}$ as well as
  $\vec{B}$ and $f_s$ may be taken to be shearing periodic
  \citep{Lees1972,Hawley1995}.}

@@

From now on we will assume no stratification ($\vfreq=0$) and axisymmetric
disturbances ($\partial/\partial{}y=0$). The dynamical equations then have
constant coefficients and we may assume that all disturbances depend on
position $\vec{r}$ and time $t$ as $\exp(ik_x{}x+ik_z{}z-i\omega{}t)$. For
such disturbances, Faraday's law \labelcref{eq:faraday-relative} is easily
solved for $\delta\vec{B}$, to wit
\begin{equation}
  \label{eq:linearized-faraday}
  \delta\vec{B} = \frac{\vec{k}}{\omega}\times\left[
    \left(\mathbf{1} - \frac{q\Omega}{i\omega}\ex\ey\right)
    \cdot\delta\tilde{\vec{E}}
  \right].
\end{equation}
Here, the second term in parentheses arises from the stretching term in
\cref{eq:faraday-relative}. Substituting \cref{eq:linearized-faraday} into
\cref{eq:formal-solution}, taking the first moment, multiplying by $\qs{}n_s$,
and summing over species yields the perturbed current density
\begin{equation}
  \label{eq:perturbed-current}
  \delta\vec{J}(\vec{r},t)
  = -\frac{1}{i\omega}\sums\frac{\qs^2n_s}{m_s}
  \int d^3v\,\int_{-\infty}^t dt'
  \tvec{v}\pder{\hat{f}_s}{\tvec{v}'}\cdot\big[
  i(\omega - \vec{k}\cdot\tvec{v}')\mathbf{1}
  + i\vec{k}\tvec{v}' + q\Omega\ex\ey
  \big]\cdot\left(\mathbf{1} - \frac{q\Omega}{i\omega}\ex\ey\right)
  \cdot\delta\tvec{E}(\vec{r}',t'),
\end{equation}
where $\hat{f}_s=f_s/n_s$ is the reduced distribution function.

@@

We are now in a position to write down the general dispersion relation of a
charge-neutral collisionless plasma in the shearing sheet approximation.
Inserting the solution of Faraday's law given in \cref{eq:linearized-faraday}
into Ampère's law $\mu_0\delta\vec{J}=i\vec{k}\times\delta\vec{B}$ (neglecting
the displacement current), and eliminating the perturbed current using
\cref{eq:conductive-relation} yields
\begin{equation}
  \label{eq:dispersion-relation}
  \mathbf{D}\cdot\left(
    \mathbf{1} - \frac{q\Omega}{i\omega}\ex\ey
  \right)\cdot\delta\tvec{E} = 0,
\end{equation}
where the dispersion tensor
\begin{equation}
  \label{eq:dispersion-tensor}
  \mathbf{D} =
  (k^2\mathbf{1} - \vec{k}\vec{k} - i\omega\mu_0\vec{\sigma})\va^2
  - 2q\Omega^2\sums\frac{n_s m_s}{\rho}
  \frac{\omcs/b_z}{\omcs b_z + 2\Omega}\ey\ey.
\end{equation}
Here, $\rho=\sums{}n_s{}m_s$ is the mass density, $\va^2=B^2/\mu_0\rho$ is
the square of the Alfvén speed, and we have used the charge-neutrality
condition to rewrite the last term on the right hand side.

@@

Here we consider the stability of an initially unmagnetized cold plasma in the
shearing sheet. The dispersion relation for this case is obtained from the
general expression \cref{eq:dispersion-tensor} together with
\cref{eq:conductivity} by substituting \cref{eq:cold-response} and then
letting the magnetic field strength tend to zero. The dispersion tensor in
this limit reduces to
\begin{equation}
  \label{eq:unmagnetized-dispersion-tensor}
  \va^{-2}\ell^2\mathbf{D} =
  (k^2\mathbf{1} - \vec{k}\vec{k} - i\omega\mu_0\vec{\sigma})\ell^2
  + q/2\,\ey\ey,
\end{equation}
where $\ell$ is the mean inertial length, defined through
\begin{equation}
  \ell^{-2} = \mu_0\sums\frac{\qs^2n_s}{m_s}.
\end{equation}
The limiting form of the cold plasma conductivity tensor is
\begin{equation}
  \label{eq:zero-mag-conductivity}
  -i\omega\mu_0\vec{\sigma} =
  \ell^{-2}\mathbf{Q}\cdot\mathbf{\Lambda}\cdot\mathbf{Q},
\end{equation}
where the anisotropy tensor is given in \cref{eq:anisotropy-tensor} with
$\Delta_s=q/2$ and the cold plasma response tensor for $B=0$ is
\begin{equation}
  \label{eq:zero-mag-response}
  \mathbf{\Lambda} = \ez\ez
  + \frac{1}{2}\sum_{\pm}\frac{\omega}{\omega\pm\kappa}
  (\ex\pm i\ey)(\ex\mp i\ey).
\end{equation}
With this, setting the determinant of \cref{eq:unmagnetized-dispersion-tensor}
equal to zero yields the dispersion relation
\begin{equation}
  \frac{\omega^2{(1 + k^2\ell^2)}^2 - \kappa^2 k_z^2\ell^2(q/2 + k^2\ell^2)}
  {\omega^2 - \kappa^2} = 0.
\end{equation}
Assuming $\omega^2\ne\kappa^2$, the solution is
\begin{equation}
  \omega^2 =
  \frac{\kappa^2k_z^2\ell^2(q/2 + k^2\ell^2)}{{(1 + k^2\ell^2)}^2}.
\end{equation}
This is manifestly positive as long as $q>0$. Thus there is no instability as
long as the orbital frequency decreases radially outward.

@@

In most plasmas, electrons are much lighter than ions, and so the mass ratio
$m_e/m_i\ll1$. If we assume that the electrons are also much colder than the
ions ($T_e/T_i\ll1$), then the electrons can be modeled as a cold massless
fluid. Expanding the cold electron conductivity tensor in the mass ratio
yields
\begin{equation}
  \label{eq:cold-electron-conductivity}
  \vec{\sigma}_e \approx -\frac{1}{i\omega}\frac{e^2n_e}{m_e}\vec{b}\vec{b}
  + \frac{en_e}{B}\vec{b}\times\mathbf{1},
\end{equation}
where we have dropped terms of order $m_e/m_i$ and higher. The first term
evidently diverges as the mass ratio goes to zero. Because this term is
proportional to $\vec{b}\vec{b}$, the dispersion tensor becomes
block-diagonal.\footnote{We note that since
  $\mathbf{D}=-i\omega\mu_0\vec{\sigma}_e\va^2+\ldots$, it follows from the
  dispersion relation given in \cref{eq:dispersion-relation} that the parallel
  component of the (relative) electric field must be at least first order in
  $m_e/m_i$. The parallel current density would otherwise diverge, a result
  that is consistent with guiding center theory \citep[e.g.][]{Grad1961}.} In
the cold and massless electron limit, the dispersion relation thus factorizes
into $\delta{}\tilde{E}_\parallel=0$ and $\det\mathbf{D}_\perp=0$, where the
perpendicular dispersion tensor
\begin{equation}
  \label{eq:perpendicular-dispersion}
  \mathbf{D}_\perp =
  {(k^2\mathbf{1} - \vec{k}\vec{k} - i\omega\mu_0\vec{\sigma}_i)}_\perp\va^2
  - i\omega\ome\vec{b}\times\mathbf{1}
  - 2q\Omega^2\sums\frac{n_s{}m_s}{\rho}\frac{\omcs b_z}{\omcs b_z + 2\Omega}
  (\vec{b}\times\ex)(\vec{b}\times\ex).
\end{equation}
We note that throughout this section, the labels $\perp$ and $\parallel$ refer
to the direction of $\vec{b}$, \emph{not} to the gyro-frequency vector $\vomg$
as in \cref{sec:steady-state,sec:linear-theory}. In
\cref{eq:perpendicular-dispersion}, $\vec{\sigma}_i$ is the (multi-component)
ion conductivity tensor and
\begin{equation}
  \ome = \frac{en_e B}{\rho} = \sums\frac{n_s m_s}{\rho}\omcs
\end{equation}
is the (mass density weighted) mean ion cyclotron frequency.

@@

In the remainder of this section we will assume that there is only one ion
species. The perpendicular dispersion tensor in this case is given by
\begin{equation}
  \label{eq:perpendicular-dispersion-single}
  \mathbf{D}_\perp =
  \va^2{(k^2\mathbf{1} - \vec{k}\vec{k})}_\perp
  + \omc^2{(\mathbf{Q}\cdot\mathbf{\Lambda}\cdot\mathbf{Q})}_\perp
  - i\omega\omc\vec{b}\times\mathbf{1}
  - 2q\Omega^2\omc b_z/S_z
  (\vec{b}\times\ex)(\vec{b}\times\ex).
\end{equation}
We focus on the Vlasov-fluid limit in this section to make contact with the
existing kinetic theory calculations of the MRI, all of which have assumed
cold electrons. We stress, however, that the results derived in the previous
section are fully general and can be used to quantify the effects of kinetic
electrons on the MRI\@. The cold electron approximation used here and in
previous studies formally requires that $\zeta_n\rightarrow\infty$ for all $n$
(\cref{sec:cold}). However,
$\zeta_0=\omega/|k_\parallel|v_{te}\sim{}\va/v_{te}$ for the MRI (taking
$\omega\sim{}k_\parallel\va$). Thus $\zeta_0\gg{}1$ requires
$T_e/T_i\ll(m_e/m_i)\beta_i^{-1}$. This is in general \emph{not} satisfied so
that the cold electron approximation is not formally applicable when studying
the MRI in hot accretion flows. In future work, we will study the effect of
(hot) kinetic electrons on the MRI in collisionless plasmas.

@@

In the limit $e/m\to\infty$ we should be able to recover the results of
\citet{Quataert2002} who analyzed the stability of a differentially rotating
plasma in the shearing sheet approximation using the so-called kinetic MHD
equations \citep[see e.g.][]{Grad1961,Kulsrud1983,Hazeltine2004}. We start by
expanding the response tensor in powers of $m/e$. Keeping mostly leading order
terms, this yields
\begin{equation}
  \mathbf{\Lambda} \approx
  -\frac{\omega^2}{\omc^2}(\mathbf{1} - \vec{e}_3\vec{e}_3)
  + \frac{i\omega}{\omc}\left[1 - \frac{(4-q)\Omega b_z}{2\omc}\right]
  \vec{e}_3\times\mathbf{1}
  - \frac{2(W-1)k_\perp^2\vt^2}{\omc^2}\vec{e}_2\vec{e}_2
  + \frac{i\zeta_0 Wk_\perp\vt}{\omc}\vec{e}_1\times\mathbf{1}
  - \zeta_0^2 W\vec{e}_3\vec{e}_3.
\end{equation}
The argument of the $W$-function is $\zeta_0=\omega/|k_\parallel|\vt$.
Expanding the anisotropy tensor $\mathbf{Q}$ yields
\begin{equation}
  \mathbf{Q} \approx \mathbf{1} - \frac{q\Omega}{2\omc b_z}
  \left[1 - \frac{(8-q)\Omega}{4\omc b_z}\right]\ey\ey.
\end{equation}

@@

In this section we study modes whose wave vector is aligned with the magnetic
field, which in turn is aligned with the rotation axis. The ion response
tensor in this case is given by
\begin{equation}
  \label{eq:parallel-response}
  \mathbf{\Lambda} =
  \frac{1}{2}\sum_\pm\Lambda^{}_\pm
  (\vec{e}_1\pm i\vec{e}_2)(\vec{e}_1\mp i\vec{e}_2)
  - \frac{\omega^2}{k_z^2\vt^2} W\left(\frac{\omega}{|k_z|\vt}\right)
  \vec{e}_3\vec{e}_3,
\end{equation}
with perpendicular components
\begin{equation}
  \label{eq:circular-components}
  \Lambda_\pm = \frac{\omega}{\omega\pm\omg}
  \left[1 - W\left(\frac{\omega\pm\omg}{|k_z|\vt}\right)\right].
\end{equation}
We note that $\ex\ne\vec{e}_1$ and $\ey\ne\vec{e}_2$ in general even for
parallel modes. This is because the orthonormal basis
$(\vec{e}_1,\vec{e}_2,\vec{e}_3)$ with $\vec{e}_3=\vomg/\omg$ is taken to be
right-handed (see \cref{sec:plasma-response}), and whether or not the
gyro-frequency vector $\vomg$ defined in \cref{eq:gyro-frequency-vector} is
aligned with the angular velocity vector $\vec{\Omega}=\Omega\ez$ depends on
the sign of $S_z=\omc{}b_z+2\Omega$. In this context we also note that the
gyration frequency $\omg>0$ throughout this text.\footnote{Unless of course
  $\omg^2<0$, in which case much of our analysis is invalidated, see the
  discussion in \cref{sec:delta}.} The cyclotron frequency $\omc=eB/m$ can
have either sign, but this depends only on the sign of the charge, \emph{not}
on the direction of the magnetic field.

@@

We seek an axisymmetric equilibrium with a purely rotational bulk flow common
to all plasma species. In a cylindrical coordinate system $(r,\varphi,z)$,
this bulk flow may be written as
\begin{equation}
  \vec{u} = r\Omega\nabla\varphi,
\end{equation}
where $\Omega(r,z)$ is the orbital frequency. Following
\citet{Kaufman1972,Cary1983} we write the magnetic field as
\begin{equation}
  \label{eq:global-bfield}
  \vec{B} =
  \nabla w(\psi)\times\nabla\vartheta - \nabla\psi\times\nabla\varphi,
\end{equation}
where $\psi(r,z)$ is the poloidal flux function. The poloidal angle
$\vartheta(r,z)$ may be chosen to increase clockwise from $-\pi$ to $\pi$
along poloidal field lines, which are isocontours of $\psi$. Note that the
first term on the right hand side of \cref{eq:global-bfield} represents the
toroidal field. The second term is the usual parametrization of an
axisymmetric poloidal field. The magnetic field given in
\cref{eq:global-bfield} derives from the axisymmetric vector potential
\begin{equation}
  \vec{A} = w(\psi)\nabla\vartheta - \psi\nabla\varphi.
\end{equation}
